SHARP - II: Revealing a bias in observational measurements of dark matter substructure with gravitational lens flux ratios

### 2015-12-07 papers/B1555_EdgeOn/Resubmission/referee_response.txt

> 
> The final section of the paper speculates that massive disc components
> may be prevalent in strong lenses, and that such discs may bias
> inferences on the abundance of dark matter substructure using strong
> lensing.  The authors fail to provide evidence supporting these
> speculations, even though they have the data to do so.  As noted
> above, strong lenses with extended source emission (arcs and rings)
> tightly constrain the presence of thin components in lens galaxies.
> In some cases with Einstein rings, disc components are excluded from
> contributing more than percent-level fractions of the mass density.
> It would therefore be highly surprising for a substantial fraction of
> strong lenses to contain hidden, massive edge-on discs that could
> distort the lensing so significantly, as the paper intimates.
> 
> Rather than wildly speculating, the authors could easily check the
> plausibility of this claim.  A number of the authors were members of
> the SLACS collaboration, whose dataset contains many lenses with
> extended emission from the source galaxies.  These systems are the
> obvious way to quantify the fraction of lens systems containing
> massive edge-on discs, since the arcs in many SLACS lenses span a wide
> range of azimuthal angle.  Once they have quantified the abundance of
> massive edge-on discs, the authors should then perform Monte Carlo
> calculations of quasar lenses to calculate the statistics of flux
> ratio anomalies produced by these discs.  Only then can they state
> whether hidden edge-on massive discs can bias observational
> measurements of dark matter substructure with gravitational lens flux
> ratios.

RESPONSE: In the SHARP sample, we find that 5 out of the 25 systems
that we imaged have clear disk components, indicating that this is not
a completely unexpected phenomenon.  While we agree with the referee
that it is important to quantify how common disks are in lensing
galaxies, we do not feel that the SLACS sample provides the
appropriate comparison.  This is because a significant portion of the
SHARP systems come from source-selected samples, such as CLASS, that
are agnostic about the nature of the lensing galaxy.  In fact, CLASS
contains at least one other system, B1600+434 (albeit a double), where
the lensing galaxy appears to be an edge-on disk with a very low
bulge/disk ratio.  In contrast, SLACS lenses were selected to be
systems where the lensing galaxy was known a priori to be a massive
elliptical galaxy.  We would therefore expect a lower incidence of
disky systems in SLACS than in the SHARP sample.  The SWELLS sample
shows a number of massive disc systems, that contribute significantly
to the mass within 2.2 scale radii [WE DON'T MENTION THIS IN THE TEXT.
SHOULD WE, AS LEON SUGGESTED?]
We agree completely
with the referee that doing a quantification of the statistics of
massive disks and their effects on flux-ratio anomalies is an
extremely important task and, in fact, this is one of the future
projects that we are planning.  However, we feel that such an
extensive analysis is beyond the scope of this letter.  As such, we
have modified the text to focus more on this individual lens system
and put more caveats on the possible generalization of the results.

### 2015-12-09 papers/B1555_EdgeOn/Resubmission/referee_response.txt

> 
> The final section of the paper speculates that massive disc components
> may be prevalent in strong lenses, and that such discs may bias
> inferences on the abundance of dark matter substructure using strong
> lensing.  The authors fail to provide evidence supporting these
> speculations, even though they have the data to do so.  As noted
> above, strong lenses with extended source emission (arcs and rings)
> tightly constrain the presence of thin components in lens galaxies.
> In some cases with Einstein rings, disc components are excluded from
> contributing more than percent-level fractions of the mass density.
> It would therefore be highly surprising for a substantial fraction of
> strong lenses to contain hidden, massive edge-on discs that could
> distort the lensing so significantly, as the paper intimates.
> 
> Rather than wildly speculating, the authors could easily check the
> plausibility of this claim.  A number of the authors were members of
> the SLACS collaboration, whose dataset contains many lenses with
> extended emission from the source galaxies.  These systems are the
> obvious way to quantify the fraction of lens systems containing
> massive edge-on discs, since the arcs in many SLACS lenses span a wide
> range of azimuthal angle.  Once they have quantified the abundance of
> massive edge-on discs, the authors should then perform Monte Carlo
> calculations of quasar lenses to calculate the statistics of flux
> ratio anomalies produced by these discs.  Only then can they state
> whether hidden edge-on massive discs can bias observational
> measurements of dark matter substructure with gravitational lens flux
> ratios.

RESPONSE: In the SHARP sample, we find that 5 out of the 25 systems
that we imaged have clear disk components, indicating that this is not
a completely unexpected phenomenon.  While we agree with the referee
that it is important to quantify how common disks are in lensing
galaxies, we do not feel that the SLACS sample provides the
appropriate comparison.  This is because a significant portion of the
SHARP systems come from source-selected samples, such as CLASS, that
are agnostic about the nature of the lensing galaxy.  In fact, CLASS
contains at least one other system, B1600+434 (albeit a double), where
the lensing galaxy is an edge-on disk with a very low
bulge/disk ratio.  In contrast, SLACS lenses were selected to be
systems where the lensing galaxy was known a priori to be a massive
elliptical galaxy.  We would therefore expect a lower incidence of
disky systems in SLACS than in the SHARP sample.  The SWELLS sample
shows a number of massive disc systems, that contribute significantly
to the mass within 2.2 scale radii [WE DON'T MENTION THIS IN THE TEXT.
SHOULD WE, AS LEON SUGGESTED?]
We agree completely
with the referee that doing a quantification of the statistics of
massive disks and their effects on flux-ratio anomalies is an
extremely important task and, in fact, this is one of the future
projects that we are planning.  However, we feel that such an
extensive analysis is beyond the scope of this letter.  As such, we
have modified the text to focus more on this individual lens system
and put more caveats on the possible generalization of the results.

### 2015-12-09 papers/B1555_EdgeOn/Resubmission_1208/referee_response.txt

We thank the referee for his/her comments about the paper.  We have
responded to the comments below, and modified the text accordingly.

> Comments to the Author: This paper presents an interesting new
> observational result for the strong gravitational lens B1555+375,
> along with an abundance of poorly supported conjecture.  The
> observations of B1555+375 should be published, but the authors should
> either remove their speculations, or (more usefully) provide evidence
> in support of them.
> 
> CLASS B1555+375 is one of the "poster child" gravitational lens
> systems.  It is famous for its prototypical flux ratio anomaly, which
> has been cited as evidence for dark matter substructure in this
> system.  The new result presented in this paper is near-infrared
> imaging with Keck and HST which reveals a linear structure in the lens
> galaxy emission consistent with an edge-on thin disk.  Most
> intriguingly, this structure passes near the anomalous images, and
> could explain the observed flux anomaly without the need to invoke
> dark matter substructure.  This would be an important result, and
> hence should be published.
> 
> However, the discussion in the paper lacks important quantitative
> details.  For example, the paper does not describe the total
> luminosity of the disky component, compared to the spheroidal
> component of the lens galaxy.  This is important because the lens
> model discussed in the paper requires the disk component to comprise ~
> 10% of the mass within the Einstein radius, and it is unclear if that
> large mass fraction is consistent with the photometry shown in Figure
> 1.
>

RESPONSE: The referee is correct to point out that we did not quantify
the stellar mass in the lensing galaxy.  However, without the
redshifts of the lensing galaxy, we cannot convert the photometry into
a mass estimate.  That being said, the scale radius of the disk is
larger than the Einstein ring radius.  In this regime, studies of
edge-on disk lens galaxies (e.g., Dutton et al. 2011 and 2013) show
that the baryonic component can be significant, often contributing around
40 percent of the circular velocity at 2.2 scale radii.  We have also
performed a bulge+disk fit to the lens-galaxy surface brightness
distribution.  Our fit returns a bulge component that is consistent
with zero emission.  We have added text to the paper describing this
result.  We note that the two components in our lens model can be
thought of as (1) the edge-on disk and (2) the (minimal) bulge +
dark-matter halo of the lensing galaxy.

@@

> 
> The final section of the paper speculates that massive disc components
> may be prevalent in strong lenses, and that such discs may bias
> inferences on the abundance of dark matter substructure using strong
> lensing.  The authors fail to provide evidence supporting these
> speculations, even though they have the data to do so.  As noted
> above, strong lenses with extended source emission (arcs and rings)
> tightly constrain the presence of thin components in lens galaxies.
> In some cases with Einstein rings, disc components are excluded from
> contributing more than percent-level fractions of the mass density.
> It would therefore be highly surprising for a substantial fraction of
> strong lenses to contain hidden, massive edge-on discs that could
> distort the lensing so significantly, as the paper intimates.
> 
> Rather than wildly speculating, the authors could easily check the
> plausibility of this claim.  A number of the authors were members of
> the SLACS collaboration, whose dataset contains many lenses with
> extended emission from the source galaxies.  These systems are the
> obvious way to quantify the fraction of lens systems containing
> massive edge-on discs, since the arcs in many SLACS lenses span a wide
> range of azimuthal angle.  Once they have quantified the abundance of
> massive edge-on discs, the authors should then perform Monte Carlo
> calculations of quasar lenses to calculate the statistics of flux
> ratio anomalies produced by these discs.  Only then can they state
> whether hidden edge-on massive discs can bias observational
> measurements of dark matter substructure with gravitational lens flux
> ratios.

RESPONSE: In the SHARP sample, we find that 5 out of the 25 systems
that we imaged have clear disk components, indicating that this is not
a completely unexpected phenomenon.  While we agree with the referee
that it is important to quantify how common disks are in lensing
galaxies, we do not feel that the SLACS sample provides the
appropriate comparison.  This is because a significant portion of the
SHARP systems come from source-selected samples, such as CLASS, that
are agnostic about the nature of the lensing galaxy.  In fact, CLASS
contains at least one other system, B1600+434 (albeit a double), where
the lensing galaxy is an edge-on disk with a very low bulge/disk
ratio, and at least 2 more disky lenses.  In contrast, SLACS lenses
were selected to be systems where the lensing galaxy was known to be a
luminous red galaxy, and thus likely to be an elliptical.  We would
therefore expect a lower incidence of disc systems in SLACS than in
the SHARP sample.  The SWELLS sample, on the other hand, was selected
to have edge-on disc galaxies as lenses.  Due to the selection effects
in each case, neither of these samples will be much help in
determining the statistics of disc lenses.  The CLASS sample, with 3-4
disk lenses out of 22 systems, might be more informative if it were
not so small.  Overall, we agree completely with the referee that
doing a quantification of the statistics of massive disks and their
effects on flux-ratio anomalies is an extremely important task and, in
fact, this is one of the future projects that we are planning.
However, we feel that such an extensive analysis is beyond the scope
of this letter.  As such, we have modified the text to focus more on
this individual lens system and put more caveats on the possible
generalization of the results.

### 2015-12-09 papers/B1555_EdgeOn/Resubmission_1208/B1555_EdgeOn.tex
the complete resubmit version

\begin{figure*}
\includegraphics[width=0.9\textwidth]{b1555_3panel.eps}
\caption{
{\bf (a)} High-resolution $K^\prime$-band imaging of B1555+375 with contours from the MERLIN observations (red) of \citet{Marlow99} and the VLBA observations presented here (blue) overlaid.
%
{\bf (b)} Test, for now
%
{\bf (c)} Observed radio (blue pluses) and model-predicted (red open
circles) image positions of B1555+375. The lens-plane critical curves
are shown with the blue solid line and the source-plane caustics are
shown with the red dotted line. The position of the source is at
($-0.1953$, $-0.1494$), which is marked by the red filled circle. The
centroid positions of the two mass components of the lens (a SIE and
an exponential disc) are marked by the black filled triangles.
}
\label{fig:3panel}
\end{figure*}

>>>>

\begin{figure*}
\includegraphics[width=0.9\textwidth]{b1555_3panel.eps}
\caption{
{\bf (a)} $K^\prime$-band AO imaging of B1555+375 with contours from the MERLIN observations (red) of \citet{Marlow99} and the VLBA observations presented here (blue) overlaid.
%
{\bf (b)} $K^\prime$-band AO image with the locations of the four radio components
marked with circles.  
%
{\bf (c)} Observed radio (blue pluses) and model-predicted (red open
circles) image positions of B1555+375. The lens-plane critical curves
are shown with the blue solid line and the source-plane caustics are
shown with the red dotted line. The position of the source is at
($-0.1953$, $-0.1494$), which is marked by the red filled circle. The
centroid positions of the two mass components of the lens (a SIE and
an exponential disc) are marked by the black filled triangles.
}
\label{fig:3panel}
\end{figure*}
